---
title: "Introduction to R - R Ladies Vienna"
author: "Pia Pfeiffer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)
knitr::opts_chunk$set(echo = TRUE)
```

## What is R? [https://www.r-project.org/](https://www.r-project.org/)

> R is a programming language and software environment specifically designed for statistical computing and graphics. It provides a wide variety of statistical and graphical techniques, making it a preferred tool for data analysis and visualization among statisticians, data scientists, researchers, and analysts [...] across various industries and domains.
>
> `r tufte::quote_footer('--- OpenAI. (2024). ChatGPT (3.5) [Large language model] https://chat.openai.com')`

- Firms like Google, Pfizer, Bank of America, Amazon and Microsoft use R for their statistical analyses. 

- R is open source!  freely available for anyone to use, modify, and distribute, active community

### Statistics
- Most classical statistics and much of the latest methodology is available for use with R

- A few of these are built into the base R environment, but many are supplied as packages available through the CRAN family of Internet sites (via https://CRAN.R-project.org)

- Visualization: publication-quality plots and charts
<!-- * One of R's strengths is the ease with which well-designed publication-quality plots can be produced, including mathematical symbols and formulae where needed. -->

- Reproducibility: R promotes reproducible research by enabling users to document their analyses in a script format.

## Getting Started

### Installation

-   Downloading and installing R. https://cran.r-project.org/
-   Downloading and installing RStudio. https://posit.co/downloads/

### RStudio
- RStudio is an integrated development environment (IDE) for R. 

- It includes a console, syntax-highlighting editor that supports direct code execution, as well as tools for plotting, history, debugging and workspace management.

- RStudio is available in open source and commercial editions.

- RStudio runs on the desktop (Windows, Mac, and Linux) or in a browser connected to RStudio Server or RStudio Server Pro (Debian/Ubuntu, RedHat/CentOS, and SUSE Linux).

-   Introduction to RStudio interface.

## Before we begin

### Getting help
R is a collaborative project with many contributors. Type 'contributors()' for more information and 'citation()' on how to cite R or R packages in publications.
Type 'demo()' for some demos, 'help()' or '?' for on-line help, or 'help.start()' for an HTML browser interface to help. Type 'q()' to quit R.

- Beginner's guide: https://support.posit.co/hc/en-us/articles/201141096-Getting-Started-with-R 
- Manual: https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf
- R Cookbook: https://rc2e.com/ 
- R for Data Science: https://r4ds.hadley.nz/

### Basic commands and arithmetic operations.
- Using R as a calculator
```{r basic_commands}
3 + 4
3 - 7
2 * 20
4/2
10^3
sqrt(10)
```

- Comparisons
```{r comparisons}
3 < 5
3 > 5
3 <= 5 
3 >= 5
3 == 5     # Attention: double equal sign
3 != 5
```


###  Variables and data types.

- Assignment
```{r assignment}
x <- 3
y = 4
5 -> z # works, but not recommended
```

## Data Structures
Creating and manipulating data structures.

- Atomic vectors (elements are same type)
```{r atomic_vectors}
pyth_triple <- c(x, y, z)
?c
# indexing
pyth_triple[pyth_triple >= 4]
```

- Factors (for categorical data, can be ordered, levels are unique values)
```{r factors}
x1 <- c("Dec", "Apr", "Jan", "Mar")
x2 <- c("Dec", "Apr", "Jam", "Mar")
sort(x1)
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
y1 <- factor(x1, levels = month_levels)
y2 <- factor(x2, levels = month_levels)
sort(y1)
```

- Sequences (generate regular sequences)
```{r sequences}
seq_incr <- seq(1, 5, by = 1)
seq_decr <- 10^seq(4, 0, by = -1)
```

- Matrices (2-dimensional generalization of vectors)
```{r matrices}
mat <- matrix(0, nrow = 5, ncol = 5)
diag(mat) <- c(rep(1, 2), rep(4, 3))
```

- Data Frames (matrix-like structures, can handle different data types)
```{r data_frames}
v_char <- paste0("char", seq(1, 10, 1))
v_num <- seq(0.01, 0.1, length.out = 10)
v_bool <- c(rep(TRUE, 4), rep(FALSE, 6))
df <- data.frame(col1 = v_char,
                 col2 = v_num,
                 col3 = v_bool)
df$col1[df$col3] #indexing
df$col2[df$col2 > 0.05]
```


- Lists (general form of vector, elements can be of different type)

```{r lists}
my_list <- list(matrix = mat, 
                vector = pyth_triple, 
                fact = y2, 
                bool = TRUE)
```

## Packages

-   CRAN: https://cran.r-project.org/
-   Bioconductor: https://www.bioconductor.org/ 

-   find packages that might be useful for your field: https://cran.r-project.org/web/views/

-   popular packages for Data Science are combined in 'tidyverse' https://www.tidyverse.org/

-   Installation (CRAN / Github)
```{r installation, eval = FALSE}
install.packages("tidyverse")
remotes::install_github("tidyverse/tidyverse")
```

-   get help with packages and functions

```{r load_packages}
library(tidyverse)
citation("tidyverse")
```

```{r help_packages}
help(package = "tidyverse")
```

## Data Import and Export

-   Importing data from CSV, Excel, text files.
```{r data_import}
?read.table

path1 <- file.path("cereal", "cereal_1.txt")
path2 <- file.path("cereal", "cereal_2.txt")
path3 <- file.path("cereal", "cereal_3.csv")
path4 <- file.path("cereal", "cereal_4.csv")

cereal1 <- read.table(path1, header = T)
cereal2 <- read.table(path2)
cereal3 <- read.csv(path3, header = T)
cereal4 <- read.csv2(path4, header = T)

head(cereal1) # look at data
```

-   Data cleaning and preprocessing.
-   Useful package 'janitor': https://cran.r-project.org/web/packages/janitor/index.html
```{r data_proc}
cereal_new <- cereal2[,2:4]
colnames(cereal_new) <- c("Sodium", "Sugar", "Type")

cereal_new[5,2] <- NA
mean(cereal_new[,2])
cereal_new <- na.omit(cereal_new)
mean(cereal_new[,2])

cereal_new$Desc <- paste(cereal_new$Sugar, cereal_new$Type)
```

-   Exporting data.
```{r data_exp}
?write.table

write.csv(cereal_new, file = file.path("cereal", "cereal_mod.csv"))

save(cereal_new, file = file.path("cereal", "cereal_mod.Rdata"))
load(file = file.path("cereal", "cereal_mod.Rdata"))
```

-   package for data wrangling: dplyr https://dplyr.tidyverse.org/

## Data Visualization

-   Creating and modifying plots with base R bar plots, histograms.

```{r basic_plots}
plot(df$col2, df$col2^2, 
     type = "l", 
     lty = 1, 
     col = 2,
     xlab = "x", ylab = expression(x^2))
points(df$col2, df$col2^2, 
       pch = 19)
abline(v = 0.05, lty = 2)
abline(h = 0.0025, lty = 3)
abline(coef = c(0, 0.1), col = 4)
```
```{r bar_plot}
barplot(df$col2, 
        names.arg = df$col1, 
        cex.names = 0.9,
        col = c(rep(8, 3), rep(7, 2), 1:5))
```
```{r histogram}
hist(c(1, 1, 1, 1, 2, 3, 2, 2, 3, 4, 5, 5, 5), 
     breaks = 4,
     main = "Example for histogram",
     xlab = "Numbers",
     col = 4, lty = 2)
```

-   multivariate: pairs
```{r}
pairs(cereal_new[,1:2])
```

-   Export of plots
```{r plots_export}

```

-   ggplot2 https://ggplot2.tidyverse.org/
-   Colorbrewer https://colorbrewer2.org/
```{r ggplots}
library(ggplot2)
library(RColorBrewer)
ggplot(data = df,
       aes(x = col2, y = col2^2)) + 
  geom_line() +
  geom_point(aes(col = col3)) +
  xlab("x") + ylab(expression(x^2)) +
  labs(col = "Is it right?") +
  scale_color_brewer(palette = "Set1") +
  theme_grey()
```

## Programming

-   if else
```{r if_else}
if (3 > 5) {
  print("That's true!") 
  } else {
  print("That's false")
  }
```

-   for, while

```{r loops}
for (i in 1:3){
  print(paste("count:", i))
}

j <- 0
while (j < 4){
  print(j)
  j <- j + 1
}
```

-   functions
```{r functions}
my_plot <- function(x, y, color, linetype) {
  plot(x, y, type = "l", col = color, lty = linetype)
}

my_plot(df$col2, df$col2^2, 5, 2)
```


## R Markdown

- Reporting and reproducible research
- Introduction: https://rmarkdown.rstudio.com/
